---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# eIF4F.analysis

<!-- badges: start -->

[![R-CMD-check](https://github.com/a3609640/eIF4F.analysis/workflows/R-CMD-check/badge.svg)](https://github.com/a3609640/eIF4F.analysis/actions)
[![R](https://github.com/a3609640/eIF4F.analysis/actions/workflows/r.yml/badge.svg)](https://github.com/a3609640/eIF4F.analysis/actions/workflows/r.yml)
<!-- badges: end -->



eIF4F.analysis is an R-based computational pipeline to understand function and regulation of interactions among translation initiation complex proteins across tumor types. This package provides a toolkit for analyses including copy number status, prognosis, subunit abundance and stoichiometry, gene coexpression and pathway activity, and mRNA/protein correlation, all using publicly available multi-omics data.

## Before you begin

Perform the following steps to ensure the proper operation of this package.

[System requirements]

[Install RStudio/R]

[Install dependent libraries]

[Install eIF4F analysis package]

[Set up file directories]

[Download datasets]

[Tutorials]

[Session information]

## System requirements

This project makes use of various resource-intensive R packages, which carries relatively high demands for compute, RAM, and disk I/O resources (but not for graphics resources). Nonetheless, the necessary hardware is attainable in high-end consumer-grade systems.

### Description of development systems

The following systems have been used to execute the R scripts in this project:

1.  (verified) System76 "Serval" mobile workstation

    -   Intel i7-8700k CPU
    -   64GB RAM (DDR4-3000, non-ECC)
    -   Samsung NVMe Pro SSD
    -   Pop!_OS 22.04 LTS
    -   RStudio
    -   R 4.2.1

2.  (verified) PowerSpec G460 desktop computer

    -   Intel i7-8700k CPU
    -   64GB RAM (DDR4-3200, non-ECC)
    -   Intel M.2 SATA SSD
    -   Samsung NVMe Evo+ SSD
    -   Windows 10 Pro
    -   RStudio for Windows
    -   R 4.0.3

Additional details of these environments are provided in the "Session Information" section below.

## Install RStudio/R

1.  Download & install R 4.2.1, if not already installed.
2.  Download & install RStudio, if not already installed. <https://www.rstudio.com/products/rstudio/download/>

## Install dependent libraries
The required packages will be automatically installed the first time you run eIF4F.analysis.
The following commands are useful to install dependent R packages manually.
```{r, eval = FALSE}
# use Bioconductor version 3.15 for package installation
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.15")

# install required packages
bio_pkgs <- c(
  "AnnotationDbi", "circlize",  "clusterProfiler", "ComplexHeatmap", "corrplot", 
  "data.table", "devtools", "dplyr", "EnvStats", "eulerr", "factoextra", 
  "FactoMineR", "forcats", "forestplot", "ggfortify", "ggplot2", "ggpubr", 
  "graphics", "grDevices", "grid", "limma", "missMDA", "org.Hs.eg.db", "purrr", 
  "RColorBrewer", "ReactomePA", "readr",  "readxl", "reshape2", "scales", 
  "stats", "stringr", "survival", "survivalAnalysis", "tibble", "tidyr", 
  "tidyselect")
BiocManager::install(bio_pkgs)

# load required packages
lapply(bio_pkgs, require, character.only = TRUE)
```

## Install eIF4F analysis package

You can install the development version of `eIF4F.analysis` from [GitHub](https://github.com/) and load it in the R console.

```{r, eval = FALSE}
# Install eIF4F.analysis package 
devtools::install_github("a3609640/eIF4F.analysis")

# Load eIF4F.analysis package 
library(eIF4F.analysis)
```

## Set up file directories

Open the terminal and run the following command line to clone our GitHub repository for eIF4F.analysis. The package files will be stored under home directory as `~/eIF4F.analysis`

```{bash, eval = FALSE}
$ git clone https://github.com/a3609640/eIF4F.analysis
```

Under the `~/eIF4F.analysis/Script` folder, two R scripts `Download.R` and `Analysis.R` are stored for acquiring dataset and performing analysis.  Open `Download.R` file in RStudio. The directories for input and output files are defined in `Download.R` as the following.  
``` r
# default directory for data download and output storage
data_file_directory <- "~/Downloads/EIF_data/"
output_directory <- "~/Documents/EIF_output/"
```

If the root directory paths` ~/Download/EIF_data` and `~/Documents/EIF_output` do not suit, they may be adjusted in these lines near the top of `Download.R`.  
Change the folder names to your preferred folders like the following example and save the change. 
``` r
# change the directories for data download and output storage in Download.R script
data_file_directory <- "~/Downloads/new_data_folder/"
output_directory <- "~/Documents/new_output_folder/"
```

## Download datasets
`Download.R`  downloads all needed datasets (TGCA, GTEx, CPTAC, CCLE, etc.) from URLs and unzip them. To download datasets, run the `Download.R` file in RStudio with the following command line. 
```r
source("~/eIF4F.analysis/Script/Download.R")
```
`Download.R` will (unless modified) create `~/Downloads/Test/EIF_data`, where all needed datasets (TGCA, GTEx, CPTAC, CCLE etc.) will be stored and uncompressed. After the completion of the download and unzip steps, the `EIF_data` folder contains 16 data files, with a collective size of 15GB. 

## Tutorials

`Analysis.R` contains ten exported functions in the package to initialize package and execute all analyses presented in (Wu and Wagner, 2021). The users can simply execute the following command in RStudio to get all analysis performed and results to be automatically stored under `~/Documents/EIF_output`. 
``` r
source("~/eIF4F.analysis/Script/Analysis.R")
```

## Session information

The version information of R, Linux and attached or loaded packages for developing this package is the following.

``` r
> sessioninfo::session_info()
─ Session info ─────────────────────────────────────────────────────────────────
 setting  value
 version  R version 4.2.1 (2022-06-23)
 os       Pop!_OS 22.04 LTS
 system   x86_64, linux-gnu
 ui       RStudio
 language en_US:en
 collate  en_US.UTF-8
 ctype    en_US.UTF-8
 tz       America/New_York
 date     2022-08-16
 rstudio  2022.07.1+554 Spotted Wakerobin (desktop)
 pandoc   2.18 @ /usr/lib/rstudio/bin/quarto/bin/tools/ (via rmarkdown)

─ Packages ─────────────────────────────────────────────────────────────────────
 ! package          * version  date (UTC) lib source
   abind              1.4-5    2016-07-21 [1] CRAN (R 4.2.0)
   AnnotationDbi      1.58.0   2022-04-26 [1] Bioconductor
   ape                5.6-2    2022-03-02 [1] CRAN (R 4.2.0)
   aplot              0.1.6    2022-06-03 [1] CRAN (R 4.2.0)
   assertthat         0.2.1    2019-03-21 [1] CRAN (R 4.2.0)
   backports          1.4.1    2021-12-13 [1] CRAN (R 4.2.0)
   Biobase            2.56.0   2022-04-26 [1] Bioconductor
   BiocGenerics       0.42.0   2022-04-26 [1] Bioconductor
   BiocParallel       1.30.3   2022-06-05 [1] Bioconductor
   Biostrings         2.64.0   2022-04-26 [1] Bioconductor
   bit                4.0.4    2020-08-04 [1] CRAN (R 4.2.0)
   bit64              4.0.5    2020-08-30 [1] CRAN (R 4.2.0)
   bitops             1.0-7    2021-04-24 [1] CRAN (R 4.2.0)
   blob               1.2.3    2022-04-10 [1] CRAN (R 4.2.0)
   brio               1.1.3    2021-11-30 [1] CRAN (R 4.2.0)
   broom              1.0.0    2022-07-01 [1] CRAN (R 4.2.0)
   cachem             1.0.6    2021-08-19 [1] CRAN (R 4.2.0)
   car                3.1-0    2022-06-15 [1] CRAN (R 4.2.0)
   carData            3.0-5    2022-01-06 [1] CRAN (R 4.2.0)
   cellranger         1.1.0    2016-07-27 [1] CRAN (R 4.2.0)
   checkmate          2.1.0    2022-04-21 [1] CRAN (R 4.2.0)
   circlize           0.4.15   2022-05-10 [1] CRAN (R 4.2.0)
   cli                3.3.0    2022-04-25 [1] CRAN (R 4.2.0)
   clue               0.3-61   2022-05-30 [1] CRAN (R 4.2.0)
   cluster            2.1.3    2022-03-28 [4] CRAN (R 4.1.2)
   clusterProfiler    4.4.4    2022-06-21 [1] Bioconductor
   codetools          0.2-18   2020-11-04 [4] CRAN (R 4.0.3)
   colorspace         2.0-3    2022-02-21 [1] CRAN (R 4.2.0)
   ComplexHeatmap     2.12.0   2022-04-26 [1] Bioconductor
   corrplot           0.92     2021-11-18 [1] CRAN (R 4.2.0)
   cowplot            1.1.1    2020-12-30 [1] CRAN (R 4.2.0)
   crayon             1.5.1    2022-03-26 [1] CRAN (R 4.2.0)
   data.table         1.14.2   2021-09-27 [1] CRAN (R 4.2.0)
   DBI                1.1.3    2022-06-18 [1] CRAN (R 4.2.0)
   desc               1.4.1    2022-03-06 [1] CRAN (R 4.2.0)
   digest             0.6.29   2021-12-01 [1] CRAN (R 4.2.0)
   DO.db              2.9      2022-06-07 [1] Bioconductor
   doParallel         1.0.17   2022-02-07 [1] CRAN (R 4.2.0)
   DOSE               3.22.0   2022-04-26 [1] Bioconductor
   downlit            0.4.2    2022-07-05 [1] CRAN (R 4.2.0)
   downloader         0.4      2015-07-09 [1] CRAN (R 4.2.0)
   dplyr              1.0.9    2022-04-28 [1] CRAN (R 4.2.0)
   DT                 0.23     2022-05-10 [1] CRAN (R 4.2.0)
 P eIF4F.analysis   * 0.1.0    2022-08-13 [?] load_all()
   ellipsis           0.3.2    2021-04-29 [1] CRAN (R 4.2.0)
   enrichplot         1.16.1   2022-05-15 [1] Bioconductor
   EnvStats           2.7.0    2022-03-07 [1] CRAN (R 4.2.0)
   eulerr             6.1.1    2021-09-06 [1] CRAN (R 4.2.0)
   evaluate           0.15     2022-02-18 [1] CRAN (R 4.2.0)
   factoextra         1.0.7    2020-04-01 [1] CRAN (R 4.2.0)
   FactoMineR         2.4      2020-12-11 [1] CRAN (R 4.2.0)
   fansi              1.0.3    2022-03-24 [1] CRAN (R 4.2.0)
   farver             2.1.1    2022-07-06 [1] CRAN (R 4.2.0)
   fastmap            1.1.0    2021-01-25 [1] CRAN (R 4.2.0)
   fastmatch          1.1-3    2021-07-23 [1] CRAN (R 4.2.0)
   fgsea              1.22.0   2022-04-26 [1] Bioconductor
   flashClust         1.01-2   2012-08-21 [1] CRAN (R 4.2.0)
   forcats            0.5.1    2021-01-27 [1] CRAN (R 4.2.0)
   foreach            1.5.2    2022-02-02 [1] CRAN (R 4.2.0)
   forestplot         2.0.1    2021-09-03 [1] CRAN (R 4.2.0)
   fs                 1.5.2    2021-12-08 [1] CRAN (R 4.2.0)
   generics           0.1.3    2022-07-05 [1] CRAN (R 4.2.0)
   GenomeInfoDb       1.32.2   2022-05-15 [1] Bioconductor
   GenomeInfoDbData   1.2.8    2022-06-07 [1] Bioconductor
   GetoptLong         1.0.5    2020-12-15 [1] CRAN (R 4.2.0)
   ggforce            0.3.3    2021-03-05 [1] CRAN (R 4.2.0)
   ggfortify          0.4.14   2022-01-03 [1] CRAN (R 4.2.0)
   ggfun              0.0.6    2022-04-01 [1] CRAN (R 4.2.0)
   ggplot2            3.3.6    2022-05-03 [1] CRAN (R 4.2.0)
   ggplotify          0.1.0    2021-09-02 [1] CRAN (R 4.2.0)
   ggpubr             0.4.0    2020-06-27 [1] CRAN (R 4.2.0)
   ggraph             2.0.5    2021-02-23 [1] CRAN (R 4.2.0)
   ggrepel            0.9.1    2021-01-15 [1] CRAN (R 4.2.0)
   ggsignif           0.6.3    2021-09-09 [1] CRAN (R 4.2.0)
   ggtree             3.4.1    2022-07-12 [1] Bioconductor
   GlobalOptions      0.1.2    2020-06-10 [1] CRAN (R 4.2.0)
   glue               1.6.2    2022-02-24 [1] CRAN (R 4.2.0)
   GO.db              3.15.0   2022-06-07 [1] Bioconductor
   GOSemSim           2.22.0   2022-04-26 [1] Bioconductor
   graph              1.74.0   2022-04-26 [1] Bioconductor
   graphite           1.42.0   2022-04-26 [1] Bioconductor
   graphlayouts       0.8.0    2022-01-03 [1] CRAN (R 4.2.0)
   gridExtra          2.3      2017-09-09 [1] CRAN (R 4.2.0)
   gridGraphics       0.5-1    2020-12-13 [1] CRAN (R 4.2.0)
   gtable             0.3.0    2019-03-25 [1] CRAN (R 4.2.0)
   hms                1.1.1    2021-09-26 [1] CRAN (R 4.2.0)
   htmltools          0.5.2    2021-08-25 [1] CRAN (R 4.2.0)
   htmlwidgets        1.5.4    2021-09-08 [1] CRAN (R 4.2.0)
   httr               1.4.3    2022-05-04 [1] CRAN (R 4.2.0)
   igraph             1.3.3    2022-07-15 [1] CRAN (R 4.2.0)
   IRanges            2.30.0   2022-04-26 [1] Bioconductor
   iterators          1.0.14   2022-02-05 [1] CRAN (R 4.2.0)
   jsonlite           1.8.0    2022-02-22 [1] CRAN (R 4.2.0)
   KEGGREST           1.36.3   2022-07-12 [1] Bioconductor
   km.ci              0.5-6    2022-04-06 [1] CRAN (R 4.2.0)
   KMsurv             0.1-5    2012-12-03 [1] CRAN (R 4.2.0)
   knitr              1.39     2022-04-26 [1] CRAN (R 4.2.0)
   lattice            0.20-45  2021-09-22 [4] CRAN (R 4.1.1)
   lazyeval           0.2.2    2019-03-15 [1] CRAN (R 4.2.0)
   leaps              3.1      2020-01-16 [1] CRAN (R 4.2.0)
   lifecycle          1.0.1    2021-09-24 [1] CRAN (R 4.2.0)
   limma              3.52.2   2022-06-19 [1] Bioconductor
   magrittr           2.0.3    2022-03-30 [1] CRAN (R 4.2.0)
   MASS               7.3-58   2022-07-14 [1] CRAN (R 4.2.0)
   Matrix             1.4-1    2022-03-23 [4] CRAN (R 4.1.2)
   matrixStats        0.62.0   2022-04-19 [1] CRAN (R 4.2.0)
   memoise            2.0.1    2021-11-26 [1] CRAN (R 4.2.0)
   mice               3.14.0   2021-11-24 [1] CRAN (R 4.2.0)
   missMDA            1.18     2020-12-11 [1] CRAN (R 4.2.0)
   munsell            0.5.0    2018-06-12 [1] CRAN (R 4.2.0)
   mvtnorm            1.1-3    2021-10-08 [1] CRAN (R 4.2.0)
   nlme               3.1-158  2022-06-15 [1] CRAN (R 4.2.0)
   org.Hs.eg.db       3.15.0   2022-06-07 [1] Bioconductor
   patchwork          1.1.1    2020-12-17 [1] CRAN (R 4.2.0)
   pillar             1.7.0    2022-02-01 [1] CRAN (R 4.2.0)
   pkgconfig          2.0.3    2019-09-22 [1] CRAN (R 4.2.0)
   pkgdown          * 2.0.5    2022-06-23 [1] CRAN (R 4.2.0)
   pkgload            1.3.0    2022-06-27 [1] CRAN (R 4.2.0)
   plyr               1.8.7    2022-03-24 [1] CRAN (R 4.2.0)
   png                0.1-7    2013-12-03 [1] CRAN (R 4.2.0)
   polyclip           1.10-0   2019-03-14 [1] CRAN (R 4.2.0)
   purrr              0.3.4    2020-04-17 [1] CRAN (R 4.2.0)
   qvalue             2.28.0   2022-04-26 [1] Bioconductor
   R6                 2.5.1    2021-08-19 [1] CRAN (R 4.2.0)
   rappdirs           0.3.3    2021-01-31 [1] CRAN (R 4.2.0)
   RColorBrewer       1.1-3    2022-04-03 [1] CRAN (R 4.2.0)
   Rcpp               1.0.9    2022-07-08 [1] CRAN (R 4.2.0)
   RCurl              1.98-1.7 2022-06-09 [1] CRAN (R 4.2.0)
   reactome.db        1.81.0   2022-07-15 [1] Bioconductor
   ReactomePA         1.40.0   2022-04-26 [1] Bioconductor
   readr              2.1.2    2022-01-30 [1] CRAN (R 4.2.0)
   readxl             1.4.0    2022-03-28 [1] CRAN (R 4.2.0)
   reshape2           1.4.4    2020-04-09 [1] CRAN (R 4.2.0)
   rjson              0.2.21   2022-01-09 [1] CRAN (R 4.2.0)
   rlang              1.0.4    2022-07-12 [1] CRAN (R 4.2.0)
   rmarkdown          2.14     2022-04-25 [1] CRAN (R 4.2.0)
   rprojroot          2.0.3    2022-04-02 [1] CRAN (R 4.2.0)
   RSQLite            2.2.14   2022-05-07 [1] CRAN (R 4.2.0)
   rstatix            0.7.0    2021-02-13 [1] CRAN (R 4.2.0)
   rstudioapi         0.13     2020-11-12 [1] CRAN (R 4.2.0)
   S4Vectors          0.34.0   2022-04-26 [1] Bioconductor
   scales             1.2.0    2022-04-13 [1] CRAN (R 4.2.0)
   scatterpie         0.1.7    2021-08-20 [1] CRAN (R 4.2.0)
   scatterplot3d      0.3-41   2018-03-14 [1] CRAN (R 4.2.0)
   sessioninfo        1.2.2    2021-12-06 [1] CRAN (R 4.2.0)
   shadowtext         0.1.2    2022-04-22 [1] CRAN (R 4.2.0)
   shape              1.4.6    2021-05-19 [1] CRAN (R 4.2.0)
   stringi            1.7.8    2022-07-11 [1] CRAN (R 4.2.0)
   stringr            1.4.0    2019-02-10 [1] CRAN (R 4.2.0)
   survival           3.3-1    2022-03-03 [4] CRAN (R 4.1.2)
   survivalAnalysis   0.3.0    2022-02-11 [1] CRAN (R 4.2.0)
   survminer          0.4.9    2021-03-09 [1] CRAN (R 4.2.0)
   survMisc           0.5.6    2022-04-07 [1] CRAN (R 4.2.0)
   testthat           3.1.4    2022-04-26 [1] CRAN (R 4.2.0)
   tibble             3.1.7    2022-05-03 [1] CRAN (R 4.2.0)
   tidygraph          1.2.1    2022-04-05 [1] CRAN (R 4.2.0)
   tidyr              1.2.0    2022-02-01 [1] CRAN (R 4.2.0)
   tidyselect         1.1.2    2022-02-21 [1] CRAN (R 4.2.0)
   tidytidbits        0.3.2    2022-03-16 [1] CRAN (R 4.2.0)
   tidytree           0.3.9    2022-03-04 [1] CRAN (R 4.2.0)
   treeio             1.20.1   2022-07-12 [1] Bioconductor
   tweenr             1.0.2    2021-03-23 [1] CRAN (R 4.2.0)
   tzdb               0.3.0    2022-03-28 [1] CRAN (R 4.2.0)
   utf8               1.2.2    2021-07-24 [1] CRAN (R 4.2.0)
   vctrs              0.4.1    2022-04-13 [1] CRAN (R 4.2.0)
   viridis            0.6.2    2021-10-13 [1] CRAN (R 4.2.0)
   viridisLite        0.4.0    2021-04-13 [1] CRAN (R 4.2.0)
   whisker            0.4      2019-08-28 [1] CRAN (R 4.2.0)
   withr              2.5.0    2022-03-03 [1] CRAN (R 4.2.0)
   xfun               0.31     2022-05-10 [1] CRAN (R 4.2.0)
   xml2               1.3.3    2021-11-30 [1] CRAN (R 4.2.0)
   xtable             1.8-4    2019-04-21 [1] CRAN (R 4.2.0)
   XVector            0.36.0   2022-04-26 [1] Bioconductor
   yaml               2.3.5    2022-02-21 [1] CRAN (R 4.2.0)
   yulab.utils        0.0.5    2022-06-30 [1] CRAN (R 4.2.0)
   zlibbioc           1.42.0   2022-04-26 [1] Bioconductor
   zoo                1.8-10   2022-04-15 [1] CRAN (R 4.2.0)

 [1] /home/suwu/R/x86_64-pc-linux-gnu-library/4.2
 [2] /usr/local/lib/R/site-library
 [3] /usr/lib/R/site-library
 [4] /usr/lib/R/library

 P ── Loaded and on-disk path mismatch.
────────────────────────────────────────────────────────────────────────────────
```

