---
title: "eIF4F_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eIF4F_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load eIF4F.analysis library.

```{r setup}
# install.packages("devtools")
# devtools::install_github("a3609640/eIF4F.analysis")
# load eif4F.analysis	
library(eIF4F.analysis)
```

Run the `initialize_dir()` command to create the sub-directories under `~/Documents/EIF_output` to store the output files.

```{r}
initialize_dir()
```

Load the font types, size and orientation for plotting functions.

```{r, eval = FALSE}
initialize_format()
```

Load all needed TCGA, GTEX, CPTAC and CCLE datasets download by `Download.sh` for the following analyses.

```{r, eval = FALSE}
initialize_data()
```

Perform CNV analysis.

```{r, eval = FALSE}
EIF4F_CNV_analysis()
```

Perform differential gene expression/ratio analysis.

```{r, eval = FALSE}
EIF4F_DEG_analysis()
```

Perform gene expression and survival analysis.

```{r, eval = FALSE}
EIF4F_Survival_analysis()
```

Perform PCA.

```{r, eval = FALSE}
EIF4F_PCA()
```

Perform correlating gene analysis.

```{r, eval = FALSE}
EIF4F_Corrgene_analysis()
```

Perform RNA protein expression correlation analysis.

```{r, eval = FALSE}
EIF4F_RNA_pro_correlation()
```

Perform (phospho)proteomics analysis.

```{r, eval = FALSE}
EIF4F_Proteomics_analysis()
```
